Gaia's constraints on the very massive (initially $\sim$$150\,M_\odot$) runaway star candidate VFTS682

### 2018-06-25 VFTS682_APJL.tex
sec. 2 and 3 edited

In this paper we present an analysis of the new kinematical
constraints for VFTS682 provided by \emph{Gaia} DR2 and constraints
from HST proper data  by \citet{platais:18}.   We discuss the
implications for the hypothesis that VFTS682 is a slow runaway star
ejected from R136.
% which was not part of the sample of \citet{lennon:18} because of its  spectral type.

@@

@@

                        &$\mathrm{RA}_\mathrm{R136}$&[degrees] & \phantom{-}84.6767
                     &  SIMBAD  \\        
               &$     \mathrm{DEC}_\mathrm{R136}$&[degrees] &  -69.1009
                     &  SIMBAD \\       
        &$      \delta\mathrm{RA}$  &[degrees] & \phantom{-0}0.0547                      
        &  (2, 5)
  \\        
               &$     \delta\mathrm{DEC}$  &[degrees] & \phantom{-0}0.0268 
                     &  (2, 5) \\  
                       &$  \mathrm{d_\mathrm{2d}}$  &[pc] & 29
                     &  (Besthenlehner) \\

@@

 \hline     
      \multicolumn{5}{l}{\emph{HST relative proper motion for VFTS682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.01\pm0.13$ & (4) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
                  &$v_\mathrm{2d, HST}$  &[$\kms$] & $47\pm24$ & (4)\\  
                      &$\theta^{*}$  &[degrees] &   $-35_{-51}^{+24}$   & (1,5)\\  
                  %        &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
    \hline
       \multicolumn{5}{l}{\emph{Expected proper motion if ejected from R136 at zero age}} \\
      \hline
      &$v_\mathrm{2d}$  &[$\kms$] & \SdM{$xx\pm xx$} & (Bestenlehner) \\  
      &$\theta$  &[degrees] &  $0$  & \\ 
       \hline

@@

%\subsection{ Overview of VFTS682 from previous studies \label{data:vfts683}}

The WNh5 star VFTS682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} J2000, was observed as part of the multi-epoch, spectroscopic VFTS campaign covering $\lambda$4000--7000 \citep[][]{evans:11}. 
%
%of spectral type WNh5 using multi-epoch spectra covering $\lambda$4000--7000 from the
%VFTS survey. % The spectra available covered $\lambda$4000--7000 and
% were taken at multiple epochs.
%

\citet{bestenlehner:11}  analyzed the spectra to infer the stellar
parameters and measure an extinction of $A_V=4.45\pm0.12$, implying a
luminosity of $\log_{10}(L/L_\odot) =  6.5\pm0.2$, making this one of
the brightest stars in the region. The star is unlikely to have a
close companion, because of the absence of clear radial velocity
variations \citet{bestenlehner:11}. Bayesian fits of the stellar
parameters against evolutionary tracks \citep{brott:11, kohler:15}
using the Bonsaii code \citep{schneider:17, schneider:18} provide
estimates for the age, present mass and initial mass, indicating that
this star among the most massive stars known, see
\Tabref{tab:star_param} for an overview of the parameters.

\cite{townsley:06} did not find
an X-ray point source at the position of VFTS682 in their
\emph{Chandra} survey of the region, which suggests the absence of
colliding winds that would be expected in the presence of companions
even for extreme mass ratios, given the extremely large mass of VFTS682.

Estimates of the radial velocity are complicated by the variable,
optically thick winds that is typical for emission line
stars. \citet{bestenlehner:11} estimate a mass loss rate of
$10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$, not accounting for the
possible effect of clumping.  The V-band light curve of VFTS682  shows
variations at a $\sim$10\% level on a timescale of years, which is
unusual for Wolf-Rayet stars and more typical for Luminous Blue
Variable (LBV) stars \citep{udalski:08, bestenlehner:11}. It also
shows mid-infrared excess \citep{gruendl:09}.  We therefore caution for over-interpreting the estimates for the radial velocities, but we will summarize the estimates that have been published.  \citet{bestenlehner:11}  estimate a radial velocity (RV) of  $300\pm10\kms$ using the  N{\footnotesize V} $\lambda4944$ line, which is offset from the average radial velocity of the region, which is  $270\pm10\kms$. This is consistent with, but no proof of, a runaway nature.  \cite{bressert:12} note an offset compared between the RV of the star and the nebular lines from the gas filaments in its vicinity. This is consistent with, but also no proof of, the expectation if the star was not formed in situ.  Newer measurements based on new Xshooter data give different estimates for the RV ({\color{blue} Rubio-D{\' i}ez et al. in prep}). We will therefore refrain from using the RV measurements in this work.

%of $137.8^{+27.5}_ {-15.9}\Msun$, an apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass of $150.0^{+28.7}_{-17.4}\Msun$.  Further parameters of are summarized in Table~\ref{X}.
VFTS682 is relatively isolated in \emph{Spitzer} images, except for the extended
source S10 interpreted as star formation triggered by the wind of
VFTS682 itself by \cite{walborn:13}.

@@

For the distance we adopt the distance to the LMC of $50$\,kpc. The error on the distance determination is small ($\lesssim2\%$, \citealt{pietrzynski:13}) and any possible offset between  R136 or VFTS682  and the distance we adopted for the LMC is probably much smaller (likely $\ll 1\%$, e.g. \citealt{Luks+1992}). These uncertainties are negligible compared to the errors in the proper motion discussed below.

@@

VFTS682 is identified with the source id 4657685637907503744 in the
\emph{Gaia} DR2 catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}
  as a 15.65 mag star in the G band
\citep{gaia:16,brown:18}.   The number of visibility periods,
i.e. groups of observations separated from other groups by a gap of at
least four days, used in the astrometric solution is seventeen for this
star. The reported astrometric excess noise is zero.  These values
suggest that the \emph{Gaia} DR2  data for VFTS682 are
reliable. Gaia errors are correlated \citep[][]{lindengren:18}, but for this star the correlation coefficient is very small, $\rho\,(\mu_\mathrm{RA}, \mu_\mathrm{DEC})$  = 0.026, where zero is for completely un-correlated data and one is for fully correlated data. We can thus safely treat the proper motion components as uncorrelated.

@@

The 30 doradus region was target of an a two-epoch photometric campaign with HST providing observations in the F775W filter in October 2011 and October 2014 (GO-12499; P.I.: D.~J.~Lennon, \citealt{sabbi:13}). Visually, the star is clearly isolated and does not appear to have any close neighbor. This by itself gives further confidence in ability of \emph{Gaia} to measure the proper motion reliably without being affected by crowding.

 \citet{platais:15, platais:18} analyzed the HST data to determine the relative proper motion and identify candidate runaway stars. The brightest stars (V$<$14) are saturated in the data set and have been excluded from the analysis.  The very massive star VFTS682 should be one of the brightest stars in the region, the high extinction makes it faint enough such that saturation effects are not an issue.   VFTS682 is relatively slow and is therefore not discussed in their paper.  However, their measurements are useful here, since they provides an  estimate of the proper motion of VFTS 682 that is independent from the \emph{Gaia} measurement.

@@

At first glance the magnitude of the relative proper motions derived by Gaia,  $32\pm 21\kms$, and by HST, $47 \pm 24\kms$, appear consistent.  In principle they are consistent with the velocity

@@

The left panel shows the distribution in two-dimensional proper motion relative
to R136, and shows that VFTS682 is not a peculiar star in terms of
its motion relative to R136. Subtracting the mean motion of the field,
we obtain $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both indicate
a non-zero peculiar projected motion that would make it a slow runaway
star as expected by \cite{bestenlehner:11} if the star was ejected
from the cluster,  but in both cases the meaurements are also consistent with zero within 2$\sigma$. 
More importantly, both the proper motion
measurement from \emph{Gaia} and HST fall in the range expected to
reach its present day location % at $\sim$29\,pc from the cluster core
if the
star was ejected very early in its evolution. This range is
highlighted by the green vertical band, and its width is determined by
the uncertainty in the present-day age of the star (cf.~\Tabref{tab:star_param}).

The right panel shows the distributions in projected flight direction for these
stars: $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its projected relative proper motion (see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce
close to radial ejections, i.e.~$\theta\simeq0$ \todo{this is only
  true for small RV}. On this panel, the green band highlights
the range of angles corresponding to an opening angle around the
projected radial direction of 45\,degrees. The direction of motion of VFTS682 is
poorly constrained because of the large errors on the proper motion
components both in HST and \emph{Gaia} data, but
it is consistent with dynamical ejection from R136. We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VRTS682) has
large peculiar radial velocity and therefore accurate distances are
needed to contrain the flight direction in three dimensions.

@@

Assuming VFTS682 indeed originates from R136, we can calculate its kinematic
age simply as:
\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu_{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\masyr} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[corresponding to $\sim$29\,pc at LMC distance,][]{bestenlehner:11}.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is compatible with a very early
ejection from the cluster, given is apparent age.
% of $1.0\pm 0.2$\,Myr 
% \citep{schneider:18}.

Therefore, although the motion of VFTS682 is not an outlier compared
to other massive stars, it is possible to tentatively claim that it is
a slow runaway with velocity compatible with dynamical ejection from
R136 early in the evolution of both the star and the cluster.

>>>>

In this paper we present an analysis of the new kinematical
constraints for VFTS682 provided by \emph{Gaia} DR2 and constraints
from HST proper motions  by \citet{platais:18}.   We discuss the
implications for the hypothesis that VFTS682 is a slow runaway star
ejected from R136.
% which was not part of the sample of \citet{lennon:18} because of its  spectral type.

@@

The WNh5 star VFTS682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} J2000, was observed as part of the multi-epoch, spectroscopic VFTS campaign covering $\lambda$4000--7000 \citep[][]{evans:11}. 
\citet{bestenlehner:11}  analyzed the spectra to infer the stellar
parameters and measure an extinction of $A_V=4.45\pm0.12$, implying a
luminosity of $\log_{10}(L/L_\odot) =  6.5\pm0.2$, making this one of
the brightest stars in the region. The star is unlikely to have a
close companion, because of the absence of clear radial velocity
variations \citet{bestenlehner:11}. Bayesian fits of the stellar
parameters against evolutionary tracks \citep{brott:11, kohler:15}
using the BONNSAI code \citep{schneider:17, schneider:18} provide
estimates for the age, present mass and initial mass, indicating that
this star among the most massive stars known, see
\Tabref{tab:star_param}. % for an overview of the parameters.

@@

                        &$\mathrm{RA}_\mathrm{R136}$&[degrees] & \phantom{-}84.6767
                     &  SIMBAD  \\        
               &$     \mathrm{DEC}_\mathrm{R136}$&[degrees] &  -69.1009
                     &  SIMBAD \\       
        &$      \delta\mathrm{RA}$  &[marcsec] & \phantom{-0}0.0547                      
        &  (2, 5)
  \\        
               &$     \delta\mathrm{DEC}$  &[marcsec] & \phantom{-0}0.0268 
                     &  (2, 5) \\  
                       &$  \mathrm{d_\mathrm{2d}}$  &[pc] & 29
                     &  (2) \\

@@

 \hline     
      \multicolumn{5}{l}{\emph{HST relative proper motion for VFTS682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.01\pm0.13$ & (4) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
                  &$v_\mathrm{2d, HST}$  &[$\kms$] & $47\pm24$ & (4)\\  
                      &$\theta^{*}$  &[degrees] &   $-35_{-51}^{+24}$   & (1,5)\\  
                  %        &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
    \hline
       \multicolumn{5}{l}{\emph{Expected proper motion if ejected from R136 at zero age}} \\
      \hline
      &$v_\mathrm{2d}$  &[$\kms$] & $29\pm 6$ & (2) \\  
      &$\theta$  &[degrees] &  $\sim0$  & \\ 
       \hline

@@

\cite{townsley:06} did not find
an X-ray point source at the position of VFTS682 in their
\emph{Chandra} survey of the region, which suggests the absence of
colliding winds that would be expected in the presence of companions
even for extreme mass ratios, given the extremely large mass of VFTS682.
This star is also relatively isolated in \emph{Spitzer} images, except for the extended
source S10 interpreted as star formation triggered by the wind of
VFTS682 itself by \cite{walborn:13}.

Estimates of the radial velocity are complicated by the variable,
possibly inhomogeneous, optically thick wind  typical of emission line
stars. \citet{bestenlehner:11} estimate a mass loss rate of
$10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$, not accounting for the
possible effect of clumping.  The V-band light curve of VFTS682  shows
variations at a $\sim$10\% level on a timescale of years, which is
unusual for Wolf-Rayet stars and more typical for Luminous Blue
Variable (LBV) stars \citep{udalski:08, bestenlehner:11}. It also
shows mid-infrared excess \citep{gruendl:09}.  We therefore caution for over-interpreting the estimates for the radial velocities, but we will summarize the estimates that have been published.  \citet{bestenlehner:11}  estimate a radial velocity (RV) of  $300\pm10\kms$ using the  N{\footnotesize V} $\lambda4944$ line, which is offset from the average radial velocity of the region, which is  $270\pm10\kms$. This is consistent with, but no proof of, a runaway nature.  \cite{bressert:12} note an offset compared between the RV of the star and the nebular lines from the gas filaments in its vicinity. This is consistent with the expectation if the star was not formed in situ.  Newer measurements based on new X-shooter data give different estimates for the RV ({\color{blue} Rubio-D{\' i}ez et al. in prep}). We will therefore refrain from using the RV measurements in this work.

%of $137.8^{+27.5}_ {-15.9}\Msun$, an apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass of $150.0^{+28.7}_{-17.4}\Msun$.  Further parameters of are summarized in Table~\ref{X}.

@@

We adopt as distance to the LMC the value of $50$\,kpc. The error on the distance determination is small ($\lesssim2\%$, \citealt{pietrzynski:13}) and any possible offset between  R136 or VFTS682  and the distance we adopted for the LMC is probably much smaller (likely $\ll 1\%$, e.g. \citealt{Luks+1992}). These uncertainties are negligible compared to the errors in the proper motion discussed below.

@@

VFTS682 is identified with the source id 4657685637907503744 in the
\emph{Gaia} DR2 catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}
  as a 15.65 mag star in the G band
\citep{gaia:16,brown:18}.   The number of visibility periods,
i.e., groups of observations separated from other groups by a gap of at
least four days, used in the astrometric solution is seventeen for this
star. The reported astrometric excess noise is zero.  These values
suggest that the \emph{Gaia} DR2  data for VFTS682 are
reliable. % We now include the correlation coefficient in the error
          % calculation, so no need for this paragraph? Gaia errors are correlated \citep[][]{lindengren:18}, but for this star the correlation coefficient is very small, $\rho\,(\mu_\mathrm{RA}, \mu_\mathrm{DEC})$  = 0.026, where zero is for completely un-correlated data and one is for fully correlated data. We can thus safely treat the proper motion components as uncorrelated.

@@

The 30 Doradus region was target of an a two-epoch photometric campaign with HST providing observations in the F775W filter in October 2011 and October 2014 (GO-12499; P.I.: D.~J.~Lennon, \citealt{sabbi:13}). Visually, the star is clearly isolated and does not appear to have any close neighbor. This by itself gives further confidence in ability of \emph{Gaia} to measure the proper motion reliably without being affected by crowding.

\citet{platais:15, platais:18} analyzed the HST data to determine the
relative proper motion and identify candidate runaway stars. The
brightest stars (V$<$14) are saturated in the data set and have been
excluded from the analysis.  VFTS682 should be one of the brightest
stars in the region, however, the high extinction makes it faint enough such that saturation effects are not an issue.   VFTS682 is relatively slow and is therefore not discussed in their paper.  However, their measurements are useful here, since they provides an  estimate of the proper motion of VFTS682 that is independent from the \emph{Gaia} measurement.

@@

% At first glance the magnitude of the relative proper motions derived by Gaia,  $32\pm 21\kms$, and by HST, $47 \pm 24\kms$, appear consistent.  In principle they are consistent with the velocity

@@

The left panel shows the distribution in two-dimensional proper motion relative
to R136, and shows that VFTS682 is not a peculiar star in terms of
its motion relative to R136. Subtracting the mean motion of the field,
we obtain $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both values
are consistent with each other and indicate
a non-zero peculiar projected motion that would make it a slow runaway
star as expected by \cite{bestenlehner:11} if the star was ejected
from the cluster,  but in both cases the meaurements are also consistent with zero within 2$\sigma$. 
More importantly, both the proper motion
measurement from \emph{Gaia} and HST fall in the range expected to
reach its present day location % at $\sim$29\,pc from the cluster core
if the
star was ejected very early in its evolution. This range is
highlighted by the green vertical band, and its width is determined by
the uncertainty in the present-day age of the star (cf.~\Tabref{tab:star_param}).

The right panel shows the distributions in projected flight direction for these
stars: $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its projected relative proper motion (see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce
close to radial ejections, i.e.~$\theta\simeq0$. On this panel, the green band highlights
% the range of angles corresponding to
an opening angle around the
projected radial direction of 45\,degrees. The direction of motion of VFTS682 is
poorly constrained because of the large errors on the proper motion
components both in HST and \emph{Gaia} data, but
it is consistent with dynamical ejection from R136. We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VFTS682) has
large peculiar radial velocity and therefore accurate distances are
needed to contrain the flight direction in three dimensions.

@@

Assuming VFTS682 indeed originates from R136, we can calculate its kinematic
age simply as:
\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu_{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\masyr} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[corresponding to $\sim$29\,pc at LMC distance,][]{bestenlehner:11}.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is compatible with a very early
ejection from the cluster, given the apparent age of the star.
% of $1.0\pm 0.2$\,Myr 
% \citep{schneider:18}.

Therefore, although the motion of VFTS682 is not an outlier compared
to other massive stars with small proper motion errors in \emph{Gaia} DR2, it is possible to tentatively claim that it is
a slow runaway with velocity compatible with dynamical ejection from
R136 early in the evolution of both the star and the cluster.
